### 2020-12-12 paper/localpsf.tex
asdf
more intro editing

The function $\varphi_x$ is the result of applying $\mathcal{A}$ to a point source, $\delta_x$, at some fixed point $x \in \Omega$, then translating the resulting impulse response function, $\mathcal{A}\delta_x$, to re-center it at zero instead of $x$. Locally translation invariant operators act like convolution operators locally, but the convolution kernel, $\varphi_x$, varies as $x$ changes. We therefore approximate $\mathcal{A}$ by a spatially varying weighted sum of convolution operators of the form

@@

	\item The functions $w_i$ are spatially varying weighting functions that interpolate the convolution kernels $\varphi_i$.

@@

We compute one ``batch'' of $\varphi_i$'s by applying $\mathcal{A}$ to a sum of point sources, $\delta_{x_i},$ at many points $x_i$ (a ``Dirac comb''). That is, we compute
\begin{equation*}
	\eta := \sum_{i=1}^m A(~\cdot~, x_i) = \left\langle \mathcal{A}, \sum_{i=1}^m \delta_{x_i} \right\rangle^*.
\end{equation*}
Here $\sum_{i=1}^m \delta_{x_i}$ is the Dirac comb distribution, and $\left\langle \mathcal{A}, \sum_{i=1}^m \delta_{x_i} \right\rangle^*$ denotes the Riesz representation of the application of $\mathcal{A}$ to the Dirac comb. In computations, this amounts to computing $\boldsymbol{\eta} = \mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1} \left(\sum_{i=1}^m \boldsymbol{\delta}_{x_i}\right)$, where $\boldsymbol{\eta}$, $\mathbf{A}$, and $\boldsymbol{\delta}_{x_i}$ are discretized versions of $\eta$, $A$, and $\delta_{x_i}$, respectively, and $\mathbf{M}$ is a mass matrix. The batch of points $x_i$ are chosen so that the support of $\left\langle \mathcal{A}, \delta_{x_i}\right\rangle^*$ and the support of $\left\langle \mathcal{A}, \delta_{x_j}\right\rangle^*$ do not overlap (or do not overlap much) if $i \neq j$. Since the impulse responses do not overlap, we may post-process $\eta$ to recover $\varphi_i$ for all of the points $x_i$ in the batch---by applying $\mathcal{A}$ to one distribution, we recover many $\varphi_i$. This is illustrated in Figure FIG. The process is repeated to get more batches of $\varphi_i$'s, until the approximation is sufficiently accurate or a maximum number of sample points is reached. In order to choose the points $x_i$, we need to estimate the support of $\left\langle \mathcal{A}, \delta_{x_i}\right\rangle^*$ \emph{before} we compute it. The supports of the impulse responses $\left\langle \mathcal{A}, \delta_{x}\right\rangle^*$ are estimated a-priori, for all $x \in \Omega$ simultaneously, via a procedure that involves applying $\mathcal{A}^T$ to a small number of constant, linear, and quadratic functions (Section \ref{eq:mean_and_covariance_estimation}). This procedure is the only part of the paper that requires non-negativity of the integral kernel.

We choose the weighting functions, $w_i$, to be the solutions to the following optimization problems:
\begin{equation}
\label{eq:wi_optimization_problem_intro}
\begin{aligned}
\min_{w_i} &\quad \frac{1}{2}\norm{\Delta w_i}_{L^2(\Omega)}^2 \\
\text{such that} &\quad w_i(x_j) = \delta_{ij}, \quad j=1,\dots,q.
\end{aligned}
\end{equation}
Here $\Delta$ is the Laplacian on $\Omega$ with pure Neumann boundary conditions, and the right hand side of the constraint, $\delta_{ij}$, is the Kronecker delta, which takes value one if $i=j$ and zero if $i \neq j$. These are the smoothest functions, in a least-squares sense, that interpolate the points $x_i$. Computing the weighting functions $w_i$ requires solving two Poisson PDEs per weighting function, and these PDEs are solved cheaply using multigrid. We call this interpolation scheme \emph{Poisson interpolation}. The advantage of Poisson interpolation over other interpolation methods, most notably radial basis function interpolation [cite Escande etc], is that Poisson interpolation incorporates information about the domain geometry when constructing the weighting functions. Consider the Pine island glacier geometry shown in Figure REF. Regions of the glacier are separated by thin inlets of ocean. With radial basis function interpolation, points on one side of the inlet inappropriately influence the interpolation on the other side of the inlet. With Poisson interpolation, the PSF samples only influence the interpolation on their own side of the inlet.

@@

The most popular existing Hessian approximation methods are based on forming a low rank approximation of the data misfit term in the Hessian, or the data misfit term preconditioned by the prior term [CITE LOTS OF STUFF]. Either the Lanczos method or the randomized singular value decomposition are used to perform the low rank approximation using only matrix-vector products. These methods suffer from a ``data predicament''---as more informative data is used in the inversion, the numerical rank of the data misfit term in the Hessian grows, so more matrix-vector products are required to form the aforementioned low rank approximation. The ideal scenario from a scientific perspective (highly informative data) is therefore the worst case scenario from a computational perspective (large computational cost) [CITE MY DISSERTATION]. In our numerical results, we show that our operator approximation method can efficiently approximate the data misfit term in the Hessian in certain PDE constrained inverse problems. The method in this paper offers a \emph{data-scalable} alternative to conventional low-rank Hessian approximation methods, because the rank of the approximation is not limited by the number of matrix-vector products used to construct the approximation. High-rank approximations of an operator can be formed using a small number of matrix-vector products.

@@

We build upon the class of product convolution approximations in which the functions, $\varphi_i$, are impulse responses of $\mathcal{A}$ to point sources at a collection of points $x_i$ [CITE BIG LIST]. 
%The primary question in methods of this type are how to choose the points $x_i$, and how to choose the weighting functions $w_i$.
A popular choice for methods in this class is to choose the points $x_i$ to be nodes in a regular grid, and interpolate the functions $\varphi_i$ with piecewise linear interpolation [CITE NAGY], or splines [CITE SPLINES]. With a regular grid, a large number of points $x_i$ is typically required to achieve an accurate product-convolution approximation. This is computationally prohibitive in our applications, because computing the functions $\varphi_i$ requires one matrix vector product with $\mathcal{A}$ for each point $x_i$, and each of these matrix vector products is extremely computationally costly. One can reduce the computational cost by starting with a coarse grid of points $x_i$, then adaptively refining the grid only in the regions where the error in the approximation is large [CITE, CITE MY PAPER]. But even with adaptive refinement, the required number of matrix vector products with $\mathcal{A}$ can be large. The key innovation in this paper is that we pick points $x_i$ in such a way that many $\varphi_i$ are computed with each matrix vector product. We do this by taking advantage of the fact that the functions $\varphi_i$ are locally supported; existing methods typically do not require this local support property.

Our method of estimating the support of the functions $\varphi_i$ was inspired by resolution analysis in seismic imaging [CITE]. In resolution analysis, $\mathcal{A}$ is the Hessian for a seismic inverse problem, and the width of $\varphi_p$ is used to estimate of the minimum length scale on which features of the parameter can be inferred near the point $p$. If the width of $\varphi_p$ is ten meters, then features of the parameter near $p$ can be accurately inferred from data if those features are larger than roughly ten meters in size. In [CITE], the width of $\varphi_x$ is estimated to be the local autocorrelation length of the function $\mathcal{A}^T \zeta$ near $x$, where $\zeta$ is a random noise funtion. In this paper, rather than probing $\mathcal{A}^T$ with random noise functions, we use specific linear and quadratic functions. Our method estimates the support of $\varphi_p$ more accurately than resolution analysis, but our method requires that $\mathcal{A}$ has a positive integral kernel, while the resolution analysis method does not have this requirement.

@@

Matrix-vector products with product-convolution approximations can be performed using the fast Fourier transform if the problem is discretized on a regular grid [CITE]. The product-convolution approximation can then replace the original operator when using Krylov methods to solve linear systems with the original operator as the coefficient operator [CITE]. However, for complex geometries, problems are rarely discretized using regular grids, so we cannot easily use the fast Fourier transform. Furthermore, in big-data inverse problems the Hessian is typically highly ill-conditioned, so the number of Krylov iterations required is large. After constructing the product-convolution approximation, it is therefore desirable to convert the product-convolution approximation to other matrix formats that are more amenable to linear algebra operations. Wavelet compression methods have been used for this purpose [CITE WAVELETS]. Here we follow our previous work [CITE MY PAPER], in which we convert the product-convolution approximation to H-matrix format.

Conventional H-matrix construction methods require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products. There are matrix-free methods for H-matrix approximation [CITE LEXING PAPER, CITE MARTINSSON, CITE OTHERS], and these methods have been used to form H-matrix compressions of Hessians in PDE constrained inverse problems [CITE ILONA, CITE NOEMI/TUCKER HODLR]. While these matrix-free H-matrix construction methods are asymptotically scalable in theory, the required number of matrix vector products is large in practice.

@@

>>>>

The function $\varphi_x$ is the result of applying $\mathcal{A}$ to a point source, $\delta_x$, at some fixed point $x \in \Omega$, then translating the resulting impulse response function to re-center it at zero instead of $x$. Locally translation invariant operators act like convolution operators locally, but the convolution kernel, $\varphi_x$, varies as $x$ changes. We therefore approximate $\mathcal{A}$ by a spatially varying weighted sum of convolution operators of the form

@@

	\item The functions $w_i$ are spatially varying weighting functions that are used to interpolate the convolution kernels $\varphi_i$.

@@

We compute one ``batch'' of $\varphi_i$'s by applying $\mathcal{A}$ to a sum of point sources, $\delta_{x_i},$ at many points $x_i$ (a ``Dirac comb''). That is, we compute
\begin{equation*}
	\eta := \sum_{i=1}^m A(~\cdot~, x_i) = \left\langle \mathcal{A}, \sum_{i=1}^m \delta_{x_i} \right\rangle^*.
\end{equation*}
Here $\sum_{i=1}^m \delta_{x_i}$ is the Dirac comb distribution, and $\left\langle \mathcal{A}, \sum_{i=1}^m \delta_{x_i} \right\rangle^*$ denotes the Riesz representation of the application of $\mathcal{A}$ to the Dirac comb. In computations, this amounts to computing $\boldsymbol{\eta} = \mathbf{M}^{-1}\mathbf{A} \mathbf{M}^{-1} \left(\sum_{i=1}^m \boldsymbol{\delta}_{x_i}\right)$, where $\boldsymbol{\eta}$, $\mathbf{A}$, and $\boldsymbol{\delta}_{x_i}$ are discretized versions of $\eta$, $A$, and $\delta_{x_i}$, respectively, and $\mathbf{M}$ is a mass matrix. The batch of points $x_i$ are chosen so that the support of $\left\langle \mathcal{A}, \delta_{x_i}\right\rangle^*$ and the support of $\left\langle \mathcal{A}, \delta_{x_j}\right\rangle^*$ do not overlap (or do not overlap much) if $i \neq j$. Since the impulse responses do not overlap, we may post-process $\eta$ to recover $\varphi_i$ for all of the points $x_i$ in the batch---by applying $\mathcal{A}$ to one distribution, we recover many $\varphi_i$. This is illustrated in Figure FIG. The process is repeated to get more batches of $\varphi_i$'s, until the approximation is sufficiently accurate or a maximum number of sample points is reached.

In order to choose the points $x_i$, we need to estimate the support of $\left\langle \mathcal{A}, \delta_{x_i}\right\rangle^*$ \emph{before} we compute it. The supports of the functions $\left\langle \mathcal{A}, \delta_{x}\right\rangle^*$ are estimated a-priori, for all $x \in \Omega$ simultaneously, via a procedure that involves applying $\mathcal{A}^T$ to a small number of specially chosen functions (three functions if $d=1$, six if $d=2$, ten if $d=3$), then post processing the results (Section \ref{eq:mean_and_covariance_estimation}). This procedure is the only part of the paper that requires non-negativity of the integral kernel.

We choose the weighting functions, $w_i$, to be the solutions to the following optimization problems:
\begin{equation}
\label{eq:wi_optimization_problem_intro}
\begin{aligned}
\min_{w_i} &\quad \frac{1}{2}\norm{\Delta w_i}_{L^2(\Omega)}^2 \\
\text{such that} &\quad w_i(x_j) = \delta_{ij}, \quad j=1,\dots,q.
\end{aligned}
\end{equation}
Here $\Delta$ is the Laplacian on $\Omega$ with pure Neumann boundary conditions, and the right hand side of the constraint, $\delta_{ij}$, is the Kronecker delta, which takes value one if $i=j$ and zero if $i \neq j$. Since the Laplacian measures local deviation from flatness [cite Jacobsen], the functions $w_i$ are the flattest functions, in a least-squares sense, that interpolate the points $x_i$. Computing the weighting functions $w_i$ requires solving two Poisson PDEs per weighting function, and these PDEs are solved cheaply using multigrid. We call this interpolation scheme \emph{Poisson interpolation}. The advantage of Poisson interpolation over other interpolation methods, most notably radial basis function interpolation [cite Escande etc], is that Poisson interpolation incorporates information about the domain geometry when constructing the weighting functions. This is illustrated in Figure FIG.

@@

The most popular existing Hessian approximation methods are based on forming a low rank approximation of the data misfit term in the Hessian, or the data misfit term preconditioned by the prior term [CITE LOTS OF STUFF]. Either the Lanczos method or the randomized singular value decomposition [CITE RSVD] are used to perform the low rank approximation using only matrix-vector products. These methods suffer from a ``data predicament''---as more informative data is used in the inversion, the numerical rank of the data misfit term in the Hessian grows, so more matrix-vector products are required to form the aforementioned low rank approximation. The ideal scenario from a scientific perspective (highly informative data) is the worst case scenario from a computational perspective (large computational cost) [CITE MY DISSERTATION]. 
%In our numerical results, we show that our operator approximation method can efficiently approximate the data misfit term in the Hessian in certain PDE constrained inverse problems. 
The method in this paper offers a \emph{data-scalable} alternative to conventional low-rank Hessian approximation methods, because high-rank approximations of an operator can be formed using a small number of matrix-vector products.

@@

We build upon the class of product convolution approximations in which the functions, $\varphi_i$, are impulse responses of $\mathcal{A}$ to point sources at a collection of points $x_i$ [CITE BIG LIST]. 
%The primary question in methods of this type are how to choose the points $x_i$, and how to choose the weighting functions $w_i$.
A popular choice for methods in this class is to choose the points $x_i$ to be nodes in a regular grid, and interpolate the functions $\varphi_i$ with piecewise linear interpolation [CITE NAGY], or splines [CITE SPLINES]. With a regular grid, a large number of points $x_i$ is typically required to achieve an accurate product-convolution approximation. This is computationally prohibitive in our applications, because computing the functions $\varphi_i$ requires one matrix vector product with $\mathcal{A}$ for each point $x_i$, and these matrix vector products are computationally costly. One can reduce the computational cost by starting with a coarse grid of points $x_i$, then adaptively refining the grid in the regions where the error in the approximation is large [CITE, CITE MY PAPER]. But even with adaptive refinement, the required number of matrix vector products with $\mathcal{A}$ can be large. In this paper we reduce the computational cost by picking points $x_i$ such that many $\varphi_i$ are computed with each matrix vector product. We do this by taking advantage of the fact that the functions $\varphi_i$ are locally supported; existing methods typically do not require this local support property.

Our method of estimating the support of the functions $\varphi_i$ was inspired by resolution analysis in seismic imaging [CITE]. In resolution analysis, $\mathcal{A}$ is the Hessian for a seismic inverse problem, and the width of $\varphi_p$ is used to estimate of the minimum length scale on which features of the parameter can be inferred near the point $p$. If the width of $\varphi_p$ is ten meters, then features of the parameter near $p$ can be accurately inferred from data if those features are larger than roughly ten meters in size. In [CITE], the width of $\varphi_p$ is estimated to be the local autocorrelation length of the function $\mathcal{A}^T \zeta$ near $p$, where $\zeta$ is a random noise funtion. In this paper, rather than probing $\mathcal{A}^T$ with random noise functions, we use specific constant, linear, and quadratic functions. Our method estimates the support of $\varphi_p$ more accurately than resolution analysis, but our method requires that $\mathcal{A}$ has a positive integral kernel, while the resolution analysis method does not have this requirement.

@@

Matrix-vector products with product-convolution approximations can be performed using the fast Fourier transform if the problem is discretized on a regular grid [CITE]. The product-convolution approximation can then replace the original operator when using Krylov methods to solve linear systems with the original operator as the coefficient operator [CITE]. However, for complex geometries, problems are rarely discretized using regular grids, so we cannot easily use the fast Fourier transform. Furthermore, in big-data inverse problems the Hessian is typically ill-conditioned, so the required number of Krylov iterations is large. Thus, after constructing the product-convolution approximation, it is desirable to convert the product-convolution approximation to other matrix formats that are more amenable to linear algebra operations. Wavelet compression methods have been used for this purpose [CITE WAVELETS]. Here we follow our previous work [CITE MY PAPER], in which we convert the product-convolution approximation to H-matrix format.

Conventional H-matrix construction methods require access to matrix entries of the matrix being approximated, and therefore cannot be used to efficiently form H-matrix approximations of operators that are only available through matrix-vector products. There are matrix-free methods for H-matrix approximation [CITE LEXING PAPER, CITE MARTINSSON, CITE OTHERS], and these methods have been used to form H-matrix compressions of Hessians in PDE constrained inverse problems [CITE ILONA, CITE NOEMI/TUCKER HODLR]. While these matrix-free H-matrix construction methods are asymptotically scalable in theory, the required number of matrix vector products can be large in practice.

@@

Consider the Pine island glacier geometry shown in Figure REF. Regions of the glacier are separated by thin inlets of ocean. With radial basis function interpolation, points on one side of the inlet inappropriately influence the interpolation on the other side of the inlet. With Poisson interpolation, the PSF samples only influence the interpolation on their own side of the inlet.
